---
title: "questions"
author: "Felix Hemsted"
date: "30 November 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F,warning = F)
```

#Getting data

```{r dataGet}
library(tidyverse)
library(lubridate)
library(knitr)
library(ggrepel)

factors25<-read_csv("25_Portfolios_5x5.CSV",skip=15,n_max = 1107) %>% 
  rename(Date = X1) %>% 
  gather(key = "Portfolio",value = "Value",-Date) %>% 
  mutate("Size" = str_split_fixed(Portfolio," ",2)[,1],"B/M" = str_split_fixed(Portfolio," ",2)[,2])



#getting market performance and risk free rate

marketAndRiskFree<-read_csv("F-F_Research_Data_Factors.CSV",skip=3,n_max = 1108) %>% 
  rename(Date = X1,MKT = "Mkt-RF") %>% select(Date,MKT,RF) %>% 
  mutate(Date = as.numeric(Date)) %>% 
  mutate(MKT = as.numeric(MKT),RF = as.numeric(RF))

firstCorrelations<-factors25 %>% 
  filter(Date>=196401&Date<199301) %>% 
  left_join(marketAndRiskFree) %>% 
  mutate("excessReturn" = Value - RF)

groupings<-unique(factors25$Portfolio)


for(i in 1:length(groupings)){
  dataNeeded<-filter(firstCorrelations,Portfolio == groupings[i])
  
  model<-lm(formula = excessReturn~MKT,data=dataNeeded)
  dataNeeded$prediction<-predict(model,dataNeeded)
  
  modelSummary<-summary(model)
  
  beta<-modelSummary$coefficients["MKT","Estimate"]
  alpha<-modelSummary$coefficients["(Intercept)","Estimate"]
  alphaSignificance<-modelSummary$coefficients["(Intercept)","Pr(>|t|)"]
  meanReturn<-mean(dataNeeded$excessReturn)
  marketReturn<-mean(dataNeeded$MKT)
  
  
  
  r2<-modelSummary$r.squared
  tempDF<-data.frame("Size" = unique(dataNeeded$Size),"B/M" = unique(dataNeeded$`B/M`),"meanReturn" = meanReturn,"alpha" = alpha,"alphaSignificance" = alphaSignificance,"beta" = beta,"r2" = r2,"mktReturn" = marketReturn,"predictedExcessReturn" =  (beta*marketReturn),stringsAsFactors = F) 
  
  if(exists("permDF")){
    permDF<-full_join(permDF,tempDF,by=names(tempDF))
  }else{
    permDF<-tempDF
  }
}

meanReturnTable<-permDF %>% 
  mutate(Size = factor(Size,levels = c("SMALL","ME1","ME2","ME3","ME4","ME5","BIG")),
         B.M = factor(B.M,levels = c("LoBM","BM1","BM2","BM3","BM4","BM5","HiBM"))) %>% 
  select(Size,B.M,meanReturn) %>% 
  spread(key = "B.M",value = "meanReturn")

betaTable<-permDF %>% 
  mutate(Size = factor(Size,levels = c("SMALL","ME1","ME2","ME3","ME4","ME5","BIG")),
         B.M = factor(B.M,levels = c("LoBM","BM1","BM2","BM3","BM4","BM5","HiBM"))) %>% 
  select(Size,B.M,beta) %>% 
  spread(key = "B.M",value = "beta")

alphaTable<-permDF %>% 
  mutate(Size = factor(Size,levels = c("SMALL","ME1","ME2","ME3","ME4","ME5","BIG")),
         B.M = factor(B.M,levels = c("LoBM","BM1","BM2","BM3","BM4","BM5","HiBM"))) %>% 
  select(Size,B.M,alpha) %>% 
  spread(key = "B.M",value = "alpha")

alphaSignificanceTable<-permDF %>% 
  mutate(Size = factor(Size,levels = c("SMALL","ME1","ME2","ME3","ME4","ME5","BIG")),
         B.M = factor(B.M,levels = c("LoBM","BM1","BM2","BM3","BM4","BM5","HiBM"))) %>% 
  select(Size,B.M,alphaSignificance) %>% 
  spread(key = "B.M",value = "alphaSignificance")

r2<-permDF %>% 
  mutate(Size = factor(Size,levels = c("SMALL","ME1","ME2","ME3","ME4","ME5","BIG")),
         B.M = factor(B.M,levels = c("LoBM","BM1","BM2","BM3","BM4","BM5","HiBM"))) %>% 
  select(Size,B.M,r2) %>% 
  spread(key = "B.M",value = "r2")
```

#Question 1

```{r Q1}

options(digits = 2)
kable(meanReturnTable)
kable(betaTable)
kable(alphaTable)
kable(alphaSignificanceTable)
kable(r2)
```

#Question 2

Generally, returns are increasing as the Book-Market ratio increases, and as firms get smaller. This implies that the market undervalues smaller, value firms, creating the opportunity for higher returns.

#Question 3

The spread in average excess returns can be partly explained by the spread in the asset betas - or how risky each asset is relative to the market. We do see nthat the betas are slightly higher for smaller firms than for larger firms (the same case as for excess returns), but that they are smaller for firms with a higher book-market ratio - opposite what excess returns does. 

Furthermore, the increase in reurns for smaller firms is not matched 1:1 by an increase in betas for them. This suggests that investors are choosing to not buy stocks of smaller firms or those with high book-market ratios, for a reason other than the riskiness of the asset, thus depressing the price of these assets and allowing them to generate a better return than what the CAPM would predict.

#Question 4

All of the alphas highlighted in the above table are significant at the P<0.05 level. This suggests that, for certain asset portfolios, there is a statistically significant return that cannot be explained by the CAPM. The r^2^ values are fairly high, ranging from `r suppressWarnings(round(min(as.numeric(as.matrix(r2)),na.rm=T),digits=2))` at the low end to `r suppressWarnings(round(max(as.numeric(as.matrix(r2)),na.rm=T),digits=2))` at the high end. This suggests that the CAPM is fairly accurate at predicting returns, although a large amount of the excess returns observed remain unexplained.

#Question 5

```{r q5}

permDF$Portfolio = paste(permDF$Size,permDF$B.M,sep = "_")

ggplot(permDF,aes(x=predictedExcessReturn,y=meanReturn,label = Portfolio))+geom_point()+xlab("CAPM excess return")+ylab("Actual excess return")+geom_abline(slope=1,colour="#E41A1C")+geom_text_repel()
```

#Question 6

The above chart shoes that the CAPM does appear to work for a small handful of portfolios - however only around five or six at best. This reinforces the idea that the CAPM is missing some fundamental variables that create excess returns.

#Question 7

As we move from large firms to smaller firms, the excess return increases. This can be partly explained by the increase in betas observed - howver the increase does not fully explain the excess returns we see. This suggests that there are additional factors at play in addition to how an asset's price changes relative to the market.

#Question 8

With value firms, or those who have a high book-to-model ratio, we see that those portfolios with a much higher relative book value perform much better than those with a low relative book value, whilst also having much lower betas. This suggests that investors should be able to get a much higher rate of return by investing in value companies, and at a much lower risk!

This puzzle is much larger than the small / large firms puzzle, as a fundamental tenant of the CAPM is that returns should increase propotionatley with risks taken - yet this strategy allows you do to the opposite. It may be the case that investors are enamoured with the growth potential of firms with a low book-to-market ratio, and therefore underprice value firms, creating the greater returns avaliable.

#Question 9


```{r question9}
firstCorrelations1<-factors25 %>% 
  filter(Date>=19301&Date<196312) %>% 
  left_join(marketAndRiskFree) %>% 
  mutate("excessReturn" = Value - RF)

for(i in 1:length(groupings)){
  dataNeeded<-filter(firstCorrelations1,Portfolio == groupings[i])
  
  model<-lm(formula = excessReturn~MKT,data=dataNeeded)
  dataNeeded$prediction<-predict(model,dataNeeded)
  
  modelSummary<-summary(model)
  
  beta<-modelSummary$coefficients["MKT","Estimate"]
  alpha<-modelSummary$coefficients["(Intercept)","Estimate"]
  alphaSignificance<-modelSummary$coefficients["(Intercept)","Pr(>|t|)"]
  meanReturn<-mean(dataNeeded$excessReturn)
  marketReturn<-mean(dataNeeded$MKT)
  
  
  
  r2<-modelSummary$r.squared
  tempDF<-data.frame("Size" = unique(dataNeeded$Size),"B/M" = unique(dataNeeded$`B/M`),"meanReturn" = meanReturn,"alpha" = alpha,"alphaSignificance" = alphaSignificance,"beta" = beta,"r2" = r2,"mktReturn" = marketReturn,"predictedExcessReturn" =  (beta*marketReturn),stringsAsFactors = F) 
  
  if(exists("permDF1")){
    permDF1<-full_join(permDF1,tempDF,by=names(tempDF))
  }else{
    permDF1<-tempDF
  }
}

permDF1$Portfolio = paste(permDF1$Size,permDF1$B.M,sep = "_")

ggplot(permDF1,aes(x=predictedExcessReturn,y=meanReturn,label = Portfolio))+geom_point()+xlab("CAPM excess return")+ylab("Actual excess return")+geom_abline(slope=1,colour="#E41A1C")+geom_text_repel()+ggtitle("1933.1 to 1963.12")

firstCorrelations2<-factors25 %>% 
  filter(Date>=193301&Date<201501) %>% 
  left_join(marketAndRiskFree) %>% 
  mutate("excessReturn" = Value - RF)

for(i in 1:length(groupings)){
  dataNeeded<-filter(firstCorrelations2,Portfolio == groupings[i])
  
  model<-lm(formula = excessReturn~MKT,data=dataNeeded)
  dataNeeded$prediction<-predict(model,dataNeeded)
  
  modelSummary<-summary(model)
  
  beta<-modelSummary$coefficients["MKT","Estimate"]
  alpha<-modelSummary$coefficients["(Intercept)","Estimate"]
  alphaSignificance<-modelSummary$coefficients["(Intercept)","Pr(>|t|)"]
  meanReturn<-mean(dataNeeded$excessReturn)
  marketReturn<-mean(dataNeeded$MKT)
  
  
  
  r2<-modelSummary$r.squared
  tempDF<-data.frame("Size" = unique(dataNeeded$Size),"B/M" = unique(dataNeeded$`B/M`),"meanReturn" = meanReturn,"alpha" = alpha,"alphaSignificance" = alphaSignificance,"beta" = beta,"r2" = r2,"mktReturn" = marketReturn,"predictedExcessReturn" =  (beta*marketReturn),stringsAsFactors = F) 
  
  if(exists("permDF2")){
    permDF2<-full_join(permDF2,tempDF,by=names(tempDF))
  }else{
    permDF2<-tempDF
  }
}

permDF2$Portfolio = paste(permDF2$Size,permDF2$B.M,sep = "_")

ggplot(permDF2,aes(x=predictedExcessReturn,y=meanReturn,label = Portfolio))+geom_point()+xlab("CAPM excess return")+ylab("Actual excess return")+geom_abline(slope=1,colour="#E41A1C")+geom_text_repel()+ggtitle("1933.1 to 2015.1")
```

Looking at the period between January 1933 and December 1964, we can see that the CAPM fairly accurately predicts the excess returns of every portfolio, with an average r^2^ of `r round(mean(permDF1$r2),digits = 2)`, and a very small average alpha of `r round(mean(permDF1$alpha),digits=2)`. This shows that, when the CAPM was being developed in the early 1960's, it was a fairly accurate model. 

By contrast, when we look at excess returns between January 1933 and January 2015, we find that the average r^2^ of the CAPM has dropped to `r round(mean(permDF2$r2),digits=2)`, whilst the average alpha has increased to `r round(mean(permDF2$r2),digits=2)`. This must mean that, between 1964 and 2015, the CAPM has lost some of its predictive ability, as other factors have become more important in determining the returns of an asset.